A Realistic Guide to Making Data Available Alongside Code to Improve Reproducibility

### 2019-06-10 paper.Rmd
first pass at adding more about the SDSS data #80

```{r setup, echo = FALSE}
knitr::opts_chunk$set(echo = FALSE,
                      out.width = "100%")

library(kableExtra)
```

@@

Data are a fundamental currency upon which scientific discoveries are made. Without access to good data, it becomes extremely difficult, if not impossible, to advance science. Yet, a large majority of data on which published research papers are based rarely see the light of day and are only visible to the original authors [@Rowhani-Farid2016; @Stodden2018].  Sharing data sets upon publication of a research paper has benefits to individual researchers, often through increased visibility of the work [Popkin2019; @Kirk2019]. But there is also a component of benefit to the broader scientific community. This primarily comes in the form of potential for reuse in other contexts along use for training and teaching (@McKiernan2016). Assuming that the data have no privacy concerns (e.g., human subjects, locations of critically endangered species), or that the act of sharing doesn't put the authors at a competitive disadvantage, sharing data will always have a net positive benefit. First and foremost, sharing data along with other artifacts can make it easier for a reader to independently reproduce the results, thereby increasing transparency and trust in the work. The act of easy data sharing can also improves model training, as many different models can be tried and tested on latest data sources, closing the loop on research and application of statistical techniques. Existing data sets can be combined or linked with new or existing data, fostering the development and synthesis of new ideas and research areas. The biggest of these benefits is the overall increase in reproducibility.

@@

Although code sharing in science had a rocky start [@Barnes2010], more and more code writing by scientists is being shared, partly due to the rapid increase in training opportunities made available by organizations like The Carpentries, combined with the rapid adoption of Github by scientists [@Ram2013]. The bigger driver for this may also be connected to the rise in popularity of data science as a discipline distinct from statistics [@Donoho2017]. This rapid growth in data science has largely been fueled by easy access to open source software tools. Programming languages such as Python, R and Julia help scientists implement and share new methods to work with data. Each of these languages is supported by thriving communities of researchers and software developers who contribute many of the building blocks that make them popular. As of this writing, Python, Julia, and R have  167k packages [@pypi], ~ 14k packages [@cran] and ~2k  packages [@julia-pkgman] respectively. These packages form the building blocks of a data scientists daily workflow.

@@

```{r fig-effort-understanding, fig.cap = "There is a big difference in the effort to prepare data, and how easy it is to understand - look at the difference between most datasets, and something like a Randomized Control Trial (RCT)."}
library(ggplot2)

@@

One low cost and easy way to distribute data alongside compute would be to package the datasets as a data only package or as part of something larger where the methods are being implemented. The R ecosystem has many noteworthy examples of data-only packages.  One exemplar is the `nycflights13` package by Hadley Wickham [@nycflights]. This package makes available airline data for all flights departing New York city in 2013 in a tidy format, with distinct tables for metadata on airlines, airports, weather, and planes. The package not only provides ready to use binary data but also ships raw data (in a folder called `data-raw`) along with scripts used to clean them. The package was originally created as a way to provide example data to teach tidy data principles and serves as a great model for how one can publish a data package in R.

@@

Including data in a python package: [@include-data-py-pkg]
A user can typically include data (or any other arbitrary files) inside a python package by declaring it in the manifest.

@@

Another common situation that researchers face is in dealing with data that fall somewhere between small (tabular data such as CSV that are in the order of bytes to megabytes to gigabytes that can be easily compressed) to large (what falls under the umbrella of `big data`). The happy medium is generally data that are too big to fit on memory of most standard machines, but can successfully fit on disk (https://ropensci.github.io/arkdb/articles/articles/noapi.html). In this case, users who do not have the support to maintain resource intensive database servers can instead rely on light serverless databases such as MonetDB or SQLite. These databases provide disk based storage and using language agnostic interfaces, a analyst can easily query these data in manageable chunks that don't overrun memory limitations. Using software packages such as `arkdb` [@arkdb] one could easily chunk large data from flat text files to these lite databases without running into memory limitations.

@@

In **Domain Specific Venue**, data can be published in journal data papers, or venues specific to the type of data. For example, in astronomy the data is hosted at the SDSS (SDSS), Genetic data can be hosted at GenBank (@GenBank), and so on.  Some other places to put data include the open science framework [@OSF], Dataverse, [@dataverse], and Data World, [@dataworld]. Other datasets might be produced

The purpose, use, and origin of the data is an important component to consider. Data for research has a different domain compared to data collected by governments, or by journalists. Many governments or civil organizations are now making their own data available through a government website interface. Media and journalism groups are also making their data available either through places like GitHub [@pudding-data], or organizations such as Data World, or may self host their data, such as five thirty eight [@data-538].

@@

## README: A Human readable description of the data

@@

## Codebook: Human readable dictionary of data contents

@@

Table \@ref(tab:table-data-incarcerate) shows an example data dictionary table taken from the Tidy Tuesday repository on incarceration trends [@tidy-tuesday-incarcerate]. This includes information on the variable the class (type) of the variable, and a longer description of the variable.

```{r table-data-incarcerate}
dictionary_incarcerate <- data.frame(
  Variable = c("year", "urbanicity", "pop_category", "rate_per_100000"),
  Class  = c("integer (date)",
             "character",
             "character",
             "double"),
  description = c("Year",
                  "County-type (urban, suburban, small/mid, rural)",
                  "Category for population - either race, gender, or Total",
                  "Rate within a category for prison population per 100,000 people")
)

knitr::kable(dictionary_incarcerate,
             caption = "The prisoner summary data dictionary, with columns on the variable, it's class, and a short description of the contents of the variable.",
             booktabs = TRUE) %>%
  kable_styling(latex_options = c("hold_position"))
```

@@

## License: How to use and share the data

@@

### CC BY

@@

### CC0

The CC0 is a “public domain” license. Data with a CC0 license means the data owners waive all their rights to the work, and it now "owned" by the public. The data can be freely shared, this means it can be copied, modified, and distributed, even for commercial purposes _without asking permission_. When using data with CC0, it is good practice to cite the original paper, but it is not required. If you wish to use the CC0, see @choose-cc0.  For a brief overview of the CC0, see @cc0-short, and for the full license, see @cc0-long.

@@

## Citation: How you want your data to be cited

@@

## Machine readable metadata

The data dictionary provides _human readable_ information on the data. To ensure data types are preserved - dates are dates, names are characters, and so on - there needs to be some form of _machine readable_ metadata. An excellent standard for metadata is Table Schema written by frictionless data [@fritionless-table-schema]. Say for a dataset called "demographics" with three variables (Table \@ref(tab:table-machine-readable))

```{r table-machine-readable}
table_machine_readable <- data.frame(
  age = c(12L, 21L, 37L),
  height = c(161.5, 181.2, 178.3),
  nationality = c("Australian", "American", "New Zealand")
)

knitr::kable(table_machine_readable,
             caption = "Example demographics table of age, height, and nationality.",
             booktabs = TRUE) %>%
  kable_styling(latex_options = c("hold_position"))
```

@@

This contains fields such as path, and a schema with subfields name and type, for each variable. It also provides information for, licensing and features such as line breaks, and delimiters. It is built on JSON (JavaScript Object Notation), a lightweight, human-friendly, machine readable data-interchange format. Table schema is baked into formats such as csvy [@csvy], an extended `csv` format, which has additional front matter in a YAML format using Table Schema.

## Raw data: The original/first data provided

@@

## Scripts: To clean raw data ready for analysis

@@

## Analysis ready data: Final data used in analysis

@@

# Tooling for packaging data

<!-- NOTE: not sure if this quite fits here, tooling is important, but at the moment we are not sure if we should describe all tooling, as I will have a lot of examples for R, and not for Python/Julia -->

Tooling for producing these this information speeds up the process of sharing data. To help create codebooks, there are R packages such as `dataMeta`, `memisc`. and `codebook`. Codebooks are implemented in other software such as STATA, which provides a "codebook" command. Data can be packaged up in a "data package" with DataPackageR [@DataPackageR], which provides tools to wrap up data into an R package, while providing helpers with MD5sum checks that track versioning. Note that is different to [Frictionlessdata's tabular data package spec [@frictionlessdata-data-pkg].

@@

## Dataset 1: Forest Census Plot on Barro Colorado Island

Long-term field surveys, where the same type of data is repeatedly measured and collected over a long period of time, are an example of data collection where the time and financial investment would necessitate meticulously curated and easily accessible metadata. Both from the fact that the same protocol is being followed year after year, and that field data collection efforts are quite expensive, these data need to be well documented, with documentation available alongside the data.

One example of a very laborious, long-term study is the 50-hectare plot on Barro Colorado Island in Panama. Researchers at the field station have censused every single tree in a pre-defined plot 7 times over the past 30 years. More than a million trees have been counted. The data and metadata however and hard to reach. To obtain the data, one must fill out a form and agree to terms and conditions which require approval and authorship prior to publication. The biggest challenge with this study is that the data are stored on a personal FTP server of one of the authors. While the data are available in CSV and binary Rdata formats, the data storage servers do not have any metadata, README files or codebooks. A separate, public archive hosts the metadata (https://repository.si.edu/handle/10088/20925) in a PDF file (https://repository.si.edu/bitstream/handle/10088/20925/RoutputFull.pdf?sequence=1&isAllowed=y) that describe the fields.

## Dataset 2: Sensor data

@@

The SDSS provides different degrees of understanding. It is a huge project involving many people, and requiring a staggering and enormous effort to prepare. Despite this, it is still very easy to understand. This is a further reflection on the idea that high effort  can create highly understandable data, (see Figure \@ref(fig:fig-effort-understanding)). The impact of this data is high, having changed the way we understand the past, present and future of the universe itself. This impact is surely due to the care and thought put into making the data accessible and public. It is a statement of what is possible with good data sharing practices.

## Dataset 3: Most other datasets

>>>>

```{r setup, echo = FALSE}
knitr::opts_chunk$set(echo = FALSE,
                      out.width = "100%")
```

@@

Data are a fundamental currency upon which scientific discoveries are made. Without access to good data, it becomes extremely difficult, if not impossible, to advance science. Yet, a large majority of data on which published research papers are based rarely see the light of day and are only visible to the original authors (CITATION).  Sharing data sets upon publication of a research paper has benefits to individual researchers, often through increased visibility of the work. But there is also a component of benefit to the broader scientific community. This primarily comes in the form of potential for reuse in other contexts along use for training and teaching (@McKiernan2016). Assuming that the data have no privacy concerns (e.g., human subjects, locations of critically endangered species), or that the act of sharing doesn't put the authors at a competitive disadvantage, sharing data will always have a net positive benefit. First and foremost, sharing data along with other artifacts can make it easier for a reader to independently reproduce the results, thereby increasing transparency and trust in the work. The act of easy data sharing can also improves model training, as many different models can be tried and tested on latest data sources, closing the loop on research and application of statistical techniques. Existing data sets can be combined or linked with new or existing data, fostering the development and synthesis of new ideas and research areas. The biggest of these benefits is the overall increase in reproducibility.

@@

Although code sharing in science had a rocky start [@Barnes2010], more and more code writing by scientists is being shared, partly due to the rapid increase in training opportunities made available by organizations like The Carpentries, combined with the rapid adoption of Github by scientists [@Ram2013]. The bigger driver for this may also be connected to the rise in popularity of data science as a discipline distinct from statistics [@Donoho2017]. This rapid growth in data science has largely been fueled by easy access to open source software tools. Programming languages such as Python, R and Julia help scientists implement and share new methods to work with data. Each of these languages is supported by thriving communities of researchers and software developers who contribute many of the building blocks that make them popular. As of this writing, Python, Julia, and R have  167k packages (https://pypi.org/), ~ 14k packages (https://cran.r-project.org/) and ~2k  packages (https://pkg.julialang.org/) respectively. These packages form the building blocks of a data scientists daily workflow.

@@

```{r fig-effort-underrstanding, fig.cap = "There is a big difference in the effort to prepare data, and how easy it is to understand - look at the difference between most datasets, and something like a Randomized Control Trial (RCT)."}
library(ggplot2)

@@

One low cost and easy way to distribute data alongside compute would be to package the datasets as a data only package or as part of something larger where the methods are being implemented. The R ecosystem has many noteworthy examples of data-only packages.  One exemplar is the `nycflights13` package by Hadley Wickham (https://cran.r-project.org/web/packages/nycflights13/index.html). This package makes available airline data for all flights departing New York city in 2013 in a tidy format, with distinct tables for metadata on airlines, airports, weather, and planes. The package not only provides ready to use binary data but also ships raw data (in a folder called `data-raw`) along with scripts used to clean them. The package was originally created as a way to provide example data to teach tidy data principles and serves as a great model for how one can publish a data package in R.

@@

Including data in a python package: https://python-packaging.readthedocs.io/en/latest/non-code-files.html
A user can typically include data (or any other arbitrary files) inside a python package by declaring it in the manifest.

@@

Another common situation that researchers face is in dealing with data that fall somewhere between small (tabular data such as CSV that are in the order of bytes to megabytes to gigabytes that can be easily compressed) to large (what falls under the umbrella of `big data`). The happy medium is generally data that are too big to fit on memory of most standard machines, but can successfully fit on disk (https://ropensci.github.io/arkdb/articles/articles/noapi.html). In this case, users who do not have the support to maintain resource intensive database servers can instead rely on light serverless databases such as MonetDB or SQLite. These databases provide disk based storage and using language agnostic interfaces, a analyst can easily query these data in manageable chunks that don't overrun memory limitations. Using software packages such as `arkdb` (https://ropensci.github.io/arkdb/index.html) one could easily chunk large data from flat text files to these lite databases without running into memory limitations.

@@

In **Domain Specific Venue**, data can be published in journal data papers, or venues specific to the type of data. For example, in astronomy the data is hosted at the SDSS (SDSS), Genetic data can be hosted at GenBank (@GenBank), and so on.  Some other places to put data include the open science framework: https://osf.io/, Dataverse, https://dataverse.org/, and Data World, https://data.world/. Other datasets might be produced

The purpose, use, and origin of the data is an important component to consider. Data for research has a different domain compared to data collected by governments, or by journalists. Many governments or civil organizations are now making their own data available through a government website interface. Media and journalism groups are also making their data available either through places like GitHub (https://github.com/the-pudding/data), or organizations such as Data World, or may self host their data (https://data.fivethirtyeight.com/).

@@

# README: A Human readable description of the data

@@

# Codebook: Human readable dictionary of data contents

@@

Below is an example data dictionary table taken from the [Tidy Tuesday repository on incarceration trends](https://github.com/rfordatascience/tidytuesday). This includes information on the variable the class (type) of the variable, and a longer description of the variable.

__Table 1. The prisoner summary data dictionary, with columns on the variable, it's class, and a short description of the contents of the variable.__

|variable        |class     | Description |
|:---------------|:---------| :---------- |
|year            |integer (date)   | Year |
|urbanicity      |character | County-type (urban, suburban, small/mid, rural) |
|pop_category    |character | Category for population - either race, gender, or Total |
|rate_per_100000 |double    | Rate within a category for prison population per 100,000 people |

@@

# License: How to use and share the data

@@

**CC BY**

@@

**CC0**

The CC0 is a “public domain” license. Data with a CC0 license means the data owners waive all their rights to the work, and it now "owned" by the public. The data can be freely shared, this means it can be copied, modified, and distributed, even for commercial purposes _without asking permission_. When using data with CC0, it is good practice to cite the original paper, but it is not required. If you wish to use the CC0, see https://creativecommons.org/choose/zero/.  For a brief overview of the CC0, see @cc0-short, and for the full license, see @cc0-long.

@@

# Citation: How you want your data to be cited

@@

# Machine readable metadata

The data dictionary provides _human readable_ information on the data. To ensure data types are preserved - dates are dates, names are characters, and so on - there needs to be some form of _machine readable_ metadata. An excellent standard for metadata is [Table Schema](https://frictionlessdata.io/specs/table-schema/). Say for a dataset called "demographics" with three variables:

| age| height|nationality |
|---:|------:|:-----------|
|  12|  161.5|Australian  |
|  21|  181.2|American    |
|  37|  178.3|New Zealand |

_Table XX Example demographics table of age, height, and nationality._

@@

This contains fields such as path, and a schema with subfields name and type, for each variable. It also provides information for, licensing and features such as line breaks, and delimiters. It is built on JSON (JavaScript Object Notation), a lightweight, human-friendly, machine readable data-interchange format. Table schema is baked into formats such as [csvy](http://csvy.org/), an extended `csv` format, which has additional front matter in a YAML format using Table Schema.

# Raw data: The original/first data provided

@@

# Scripts: To clean raw data ready for analysis

@@

# Analysis ready data: Final data used in analysis

@@

## Tooling for packaging data

NOTE: not sure if this quite fits here, tooling is important, but at the moment we are not sure if we should describe all tooling, as I will have a lot of examples for R, and not for Python/Julia

Tooling for producing these this information speeds up the process of sharing data. To help create codebooks, there are R packages such as `dataMeta`, `memisc`. and `codebook`. Codebooks are implemented in other software such as STATA, which provides a "codebook" command. Data can be packaged up in a "data package" with [DataPackageR](https://github.com/RGLab/DataPackageR), which provides tools to wrap up data into an R package, while providing helpers with MD5sum checks that track versioning. Note that is different to [Frictionlessdata's tabular data package spec](https://frictionlessdata.io/docs/tabular-data-package/).

@@

### Dataset 1: Forest Census Plot on Barro Colorado Island

The ["Forest Ceneus Plot on Barro Colorado Island"](http://ctfs.si.edu/webatlas/datasets/bci/) provides an example of data that is time consuming to document, but also robust. It contains information on trees from a 50-hectare tree plot. Here they state:

> Censuses have been carried out in 1981-1983, 1985, 1990, 1995, 2000, 2005, 2010, and 2015. In each census, all free-standing woody stems at least 10 mm diameter at breast height were identified, tagged, and mapped. Over 350,000 individual trees have been censused over 35 years.

This is an enormous, ongoing research effort. Brief context of the data is provided at the link, along with a link to the census plot data from 2012 at [this paper](https://repository.si.edu/handle/10088/20925). This provides an abstract and description of the data, along with co-authors, and a recommended acknowledgement.

This dataset falls in the top right of the graph, at high effort to prepare, and high ease of understanding. It is high effort to prepare because it is a large research effort that not only involves expensive research in terms of time. It is high ease of understanding because

The impact of this kind of research is huge, as it provides information on biodiversity ...,  ..., ... .

### Dataset 2: Sensor data

@@

The SDSS provides different degrees of understanding. It is a huge project involving many people, and a staggering effort to prepare. Despite this, it is still very easy to understand. This is a further reflection on the idea that high effort  can create highly understandable data, (see Figure \@ref(fig:fig-effort-underrstanding)). The impact of this data is high, having changed the way we understand the past, present and future of the universe itself. This impact is surely due to the care and thought put into making the data accessible and public. It is a statement of what is possible with good data sharing practices.

**include some references to the SDSS project and some of the key/impactful findings**

As an example, the optical and spectral data are explained here: https://www.sdss.org/dr15/spectro/spectro_basics/ and can be downloaded here in special FITS dataformat, or even as a plain CSV: http://skyserver.sdss.org/dr15/en/tools/quicklook/summary.aspx?

An example plot in Figure \@ref(fig:sdss-spectra)A shows the Wavelength against Flux. Here, flux is ..., and wave is ...

```{r prepare-spectra-plot, include = FALSE}

sdss_image_path <- here::here("figures", "sdss-spectra-example-plate-1678.png")
library(magick)
library(ggplot2)
library(cowplot)

spectra <- readr::read_csv(
  here::here("data",
             "plateid-1678_mjd-53433_fiberid-425_reduction2d-v5_7_0.csv")
)

gg_spectra <- ggplot(spectra,
                     aes(x = Wavelength,
                         y = Flux)) +
  geom_line() + 
  theme_minimal() + 
  theme(aspect.ratio = 1)

plot_spectra <- ggdraw() + 
  draw_image(sdss_image_path)

```{r sdss-spectra, fig.cap = "Spectra image of wavelength against flux from SDSS. Part A shows the example image taken from SDSS website. Part B shows the replicated figure using data provided"}

plot_grid(plot_spectra, gg_spectra, labels = "AUTO")
```

A great litmus test of reproducibility for this is the fact that from downloading the CSV from the website, a similar data visualisation can be produced, seen in Figure \@ref(fig:sdss-spectra)B:

The data was downloaded from: http://dr15.sdss.org/optical/spectrum/view/data/format=csv?plateid=1678&mjd=53433&fiberid=425&reduction2d=v5_7_0

This URL contains details of the data, specifically:

* plateid = 1678
* mjd = 53433
* fiberid = 425
* reduction2d = v5.7.0

These are unique identifying features of the data, which tell us:

* plateid is the plate that they use to explore a given section of night sky. It uniquely identifies a "plug plate" used to eollcet the data
* mjd is an integer that marks the "modified julian date" of the night when observations was carried out.
* fiberid is an integer that marks the fiber number used.
* The version marked as `v5_7_0` is the pipeline version used for BOSS in DR12

### Dataset 3: Most other datasets
